---
title: "03_augment"
format: html
editor: visual
---

## Loading libraries

```{r}
#| message: false
library("tidyverse")
```

## Loading data

```{r}
first_path <- "~/projects/R_project/"
file_path <- "data/02_dat_clean.tsv" 
file_path <- file.path(first_path, file_path)
df_clean <- read_tsv(file = file_path)
```

## Augmenting data

We transform the values contained in the column "Behavior" into binary variables, so that stimulated mice are assigned 1 and non-stimulated 0 in the new column "Stimulated".

We also move to the first positions the columns that describe the conditions of the mice.

```{r}
df_aug <- df_clean |>
  mutate(Stimulated = case_when(Behavior == "C/S" ~ 1,
                                Behavior == "S/C" ~ 0)) |>
  select(MouseID, Genotype, Treatment, Stimulated, class, DYRK1A_N:CaNA_N)
```

**Creating 2 New Variables**

```{r}
df_aug1 <- df_aug |>
  mutate(MAPK_Activity = BRAF_N * pERK_N)
```

**BRAF_N**: BRAF is a protein kinase that plays a key role in the MAPK/ERK signaling pathway, which regulates cell growth, differentiation, and survival. Mutations in BRAF are commonly associated with cancers, especially melanoma.

**pERK_N**: Phosphorylated ERK (pERK) is a downstream target of the MAPK pathway and is activated by BRAF. It is involved in transmitting growth signals within cells and is essential for proliferation.

**MAPK_Activity** = BRAF_N \* pERK_N

This product could represent the pathway's activity, as pERK depends on upstream signals like BRAF for its activation.

```{r}
df_aug2 <- df_aug1 |>
  mutate(Chromatin_Activation_Index = AcetylH3K9_N / DYRK1A_N)
```

**AcetylH3K9_N**: Refers to the acetylation of lysine 9 on histone H3, a key epigenetic modification. It is associated with active chromatin and transcriptional activation, regulating gene expression during cell cycle progression and development.

**DYRK1A_N**: Represents the Dual-specificity tyrosine-regulated kinase 1A, a kinase involved in various biological processes, including neural development, synaptic function, and cell proliferation. DYRK1A is also implicated in diseases such as Down syndrome and Alzheimer’s disease

**Chromatin_Activation_Index** = AcetylH3K9_N / DYRK1A_N

This ratio could highlight the balance between kinase-driven cellular signaling (DYRK1A) and chromatin activation (AcetylH3K9).

## Saving the file

```{r}
file_path <- "data/03_dat_aug.tsv"
file_path <- file.path(first_path, file_path)


df_aug2 |>
  write_tsv(file = file_path)
```
