---
title: "03_augment"
format: html
editor: visual
---

In this file, we augment the data, simplifying the naming of the Genotype, Treatment, Behaviour variables, and adding 2 new variables to the dataset.

## Loading libraries

```{r}
#| message: false
library("tidyverse")
```

Loading data

```{r}
<<<<<<< HEAD
file_path <- "/net/pupil1/home/people/s243057/projects/group_23_project/data/02_dat_clean.csv"
df <- read.csv(file_path)
```

```{r}
num <- 1:569
ggplot(df,
       aes(x = num,
           y = radius_mean,
           color = diagnosis)) +
  geom_point()
first_path <- "~/projects/R_project/"
=======
first_path <- "/net/pupil1/home/people/s242150/projects/R_project/"
#first_path <- "~/projects/R_project/"
>>>>>>> 102d035163dc39d7dedca4709284be9d0ef84627
file_path <- "data/02_dat_clean.tsv" 
file_path <- file.path(first_path, file_path)

df_clean <- read_tsv(file = file_path)
```

## Augmenting data

We transform the values contained in the columns "Genotype", "Treatment" and "Behavior" into binary variables, so that:

-   trisomic mice are assigned 1 and control mice 0;

-   mice treated with memantine are assigned 1 and with saline 0;

-   stimulated mice are assigned 1 and non-stimulated 0 in the new column "Stimulated".

We also move to the first positions the columns that describe the conditions of the mice.

```{r}
df_aug <- df_clean |>
  mutate(Genotype = case_when(Genotype == "Ts65Dn" ~ 1,
                              Genotype == "Control" ~ 0),
         Treatment = case_when(Treatment == "Memantine" ~ 1,
                               Treatment == "Saline" ~ 0),
         Stimulated = case_when(Behavior == "C/S" ~ 1,
                                Behavior == "S/C" ~ 0)) |>
  select(MouseID, Genotype, Treatment, Stimulated, class, DYRK1A_N:CaNA_N)
```

<<<<<<< HEAD
Adding columns

```{r}
df1 <- df |>
  mutate(radius_gravity = case_when(
    radius_mean < 10 ~ "Low",
    radius_mean >= 10 & radius_mean < 20 ~ "Medium",
    radius_mean >= 20 ~ "High"),
    radius_gravity = factor(radius_gravity, levels = c("Low", "Medium", "High")))

```

```{r}
df2 <- df1 |> 
  mutate(diagnosis_binary = if_else(diagnosis == "M", 1, 0))
file_path <- "data/03_dat_aug.tsv"
file_path <- file.path(first_path, file_path)


df_aug2 |>
  write_tsv(file = file_path)
```
=======
### **Creating 2 New Variables:**
>>>>>>> 102d035163dc39d7dedca4709284be9d0ef84627

**VARIABLE 1: MAPK_Activity**

```{r}
df_aug1 <- df_aug |>
  mutate(MAPK_Activity = BRAF_N * pERK_N)
```

**BRAF_N**: BRAF (v-raf murine sarcoma viral oncogene homolog B) is a protein kinase that plays a key role in the MAPK/ERK (Mitogen-Activated Protein Kinase / Extracellular Signal-Regulated Kinase) signaling pathway, which regulates cell growth, differentiation, and survival. Mutations in BRAF are commonly associated with cancers.

**pERK_N**: Phosphorylated ERK (pERK) is a downstream target of the MAPK pathway and is activated by BRAF. It is involved in transmitting growth signals within cells and is essential for proliferation.

**MAPK_Activity** = BRAF_N \* pERK_N

This product could represent the pathway's activity, as pERK depends on upstream signals like BRAF for its activation.

**VARIABLE 2: Chromatin_Activation_Index**

```{r}
df_aug2 <- df_aug1 |>
  mutate(Chromatin_Activation_Index = AcetylH3K9_N / DYRK1A_N)
```

**AcetylH3K9_N** (acetylation of histone H3 at lysine 9): Refers to the acetylation of lysine 9 on histone H3, a key epigenetic modification. It is associated with active chromatin and transcriptional activation, regulating gene expression during cell cycle progression and development.

**DYRK1A_N** (Dual-specificity Tyrosine-phosphorylation Regulated Kinase 1A): Represents the Dual-specificity tyrosine-regulated kinase 1A, a kinase involved in various biological processes, including neural development, synaptic function, and cell proliferation. DYRK1A is also implicated in diseases such as Down syndrome and Alzheimer’s disease

**Chromatin_Activation_Index** = AcetylH3K9_N / DYRK1A_N

This ratio could highlight the balance between kinase-driven cellular signaling (DYRK1A) and chromatin activation (AcetylH3K9).

## Saving the file

```{r}
file_path <- "data/03_dat_aug.tsv"
file_path <- file.path(first_path, file_path)


df_aug2 |>
  write_tsv(file = file_path)
```
