---
title: "05_analysis_1"
format: html
editor: visual
---

# Pre-analysis

Before going into descriptive statistics we analyze the general trend of our data.

## Loading libraries

```{r}
#| message: false
library("tidyverse")
library("viridis")
```

## Loading data

```{r}
first_path <- "~/projects/R_project/"
file_path <- "data/03_dat_aug.tsv" 
file_path <- file.path(first_path, file_path)

df_aug <- read_tsv(file = file_path)
```

## Analysis

First, we count the occurance of each class in the dataset. The classes are the following:

1.  **c-CS-s**: control mice, stimulated to learn, injected with saline

2.  **c-CS-m**: control mice, stimulated to learn, injected with memantine

3.  **c-SC-s**: control mice, not stimulated to learn, injected with saline

4.  **c-SC-m**: control mice, not stimulated to learn, injected with memantine

5.  **t-CS-s**: trisomy mice, stimulated to learn, injected with saline

6.  **t-CS-m**: trisomy mice, stimulated to learn, injected with memantine

7.  **t-SC-s**: trisomy mice, not stimulated to learn, injected with saline

8.  **t-SC-m**: trisomy mice, not stimulated to learn, injected with memantine

```{r}
ggplot(data = df_aug,
       mapping = aes(class, 
                     fill = class)) +
  geom_bar(alpha = 0.8) +
  scale_fill_viridis_d(option = "D") +
  theme_light(base_family = "Avenir") +
  labs(title = "Count of each class")
```
